---
title: "BNT162b2 Vaccine Efficiency Analysis"
subtitle: "Likelihood, Bayesian, Bootstrapping"
author: "Yanting Hu, Liuyixin Shao, Shuxin Zhang"
graphics: yes
output: 
        pdf_document:
         toc: false
         number_sections: true
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{multicol,graphicx,hyperref,xcolor}
- \usepackage{setspace} \doublespacing
fontsize: 11pt

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-packages, message=FALSE, echo = F}
library(tidyverse)
library(openintro)
library(maxLik)
library(ggplot2)
library(patchwork)
```

# Abstract

The coronavirus disease 2019 (Covid-19) has afflicted tens of millions of people around the world. A total of 34,922 participants were randomly selected, while 8 out of 17,411 in the BNT162b2 group were infected and 162 out of 17,511 in the Placebo group were infected. This paper is about security and Effectiveness tests for BNT162b2 Vaccine by using three methods: Likelihood, Bayesian, and Bootstrapping. We found that BNT162b2 Vaccine conferred approximately 95% protection against Covid-19 in persons 16 years of age or beyond. The median of the vaccine's safety...

# Keywords

Covid-19; BNT162b2-vaccine; Likelihood; Bayesian; Bootstrapping

# Introduction

  In 2019, human beings experiences an unprecedented outbreak of respiratory infections. This is a worldwide disease known as coronavirus disease 2019 (Covid-19). The first cases of novel coronavirus are detected in China and then spread to other countries across the world. During more than 3 years pandemic, “over 2 million people in the European Region have died from the disease.” \footnote{(Coronavirus Disease (COVID-19) Pandemic)} At the beginning of the outbreak, people didn't expect it to last for a long time and treated it like a common flu. However, because the infected person is exposed to the crowd for a long time, the development of new coronary pneumonia is out of control and gradually forms a global disease.

  How to effectively control the epidemic? Isolation is often the first and most effective measure that comes to mind when attempting to control an epidemic. It is a logical approach since, without person-to-person contact, the virus cannot spread. However, it is essential to understand that this approach requires a sustained effort until the government can confidently ensure the quarantine of all infected individuals. It is undeniable that Quarantine and social distancing are one of the most effective ways to prevent the virus from spreading, however, this method also lead to "elevated levels of loneliness and social isolation, which in turn produce physical- and mental-health related repercussions." \footnote{(Hwang et al.)}
  
  So it becomes meaningful to control the spread of the virus from the human body itself, and vaccination is one of them. The effective vaccines can greatly reduce "diseases that once routinely harmed or killed babies, children, and adults." \footnote{(“Adult Vaccination - Reasons to Vaccinate.”)}
  
  BNT162b2 vaccine, a Covid 19 vaccine developed by Pfizer, has entered public's view. BNT162b2 is a vaccine containing modified RNA, encapsulated in lipid nanoparticles, that carries the genetic instructions for producing a stable form of the SARS-CoV-2 spike protein, which is anchored to the viral membrane. And in this research, individuals aged 16 years or older were randomly divided into two groups. One group received two doses of either a placebo or the BNT162b2 vaccine. The two doses were rigorously administered 21 days apart. And then, researchers measure the effectiveness of vaccine by record the number of occurrence of laboratory-confirmed COVID-19. By analyzing the data calculated, we could use many statistics model to illustrates the effectiveness of the BNT162b2 vaccine.
  
  Many previous studies have already explored the effects of the BNT162b2 vaccine, while this study illustrates and records a real-life experiment. In the following sections, we will analyse whether BNT162b2 vaccine is efficient and safety enough to against laboratory-confirmed Covid 19 by three methods: likelihood, Bayesian, and bootstrapping.


# Statistical Method
## Preprocessing
### Our Data
Based on the paper, we have the table:
\begin{table}[h]
\centering
\caption{Vaccine Efficacy against Covid 19 at least 7 days after second dose in patients without evidence of infection}
\begin{tabular}{cccc}
Group & Cases & No. of subjects\\ \hline
BNT162b2 & 8 & 17,411 \\
Placebo & 162 & 17,511 \\ \hline
Total & 170 & 34,922 \\ \hline
\end{tabular}
\end{table}
### Two Statistical Models and Their Corresponding Parameters and Assumptions
To do appropriate inference, our three models are built on two different approaches: 1) the two binomial proportions approach and 2) the one-sample binomial approach.

1) First, it's natural for us to treat the data into a two binomial proportions problem.

Let $X$ denote the number of infection cases among $n_1 = 17411$ subjects assigned to the BNT162b2 group and $Y$ denote the number of infection cases among $n_2 = 17511$ subjects assigned to the Placebo group.

Let $\pi_v$ and $\pi_p$ denote the probabilities of an infection in the vaccine and placebo groups respectively.

Since participants underwent randomization in the paper, every subject is independent of each other. With independence and infection probability $\pi_v$ and $\pi_p$ for the vaccine and placebo groups, a reasonable model that we can use for $X$ and $Y$ is $X \sim Binom(\pi_v, n_1 = 17411)$ and $Y \sim Binom(\pi_p, n_2 = 17511)$. 

Then, the parameter of interest, which is the vaccine efficacy is:
$\psi = 1 - \frac{\pi_v}{\pi_p} = \frac{\pi_p - \pi_v}{\pi_p}$

Based on the data from the table above, we can easily calculated that $\pi_v = \frac{8}{17411}$, $\pi_p = \frac{162}{17511}$, and $\hat{\psi} = \frac{\pi_p - \pi_v}{\pi_p}=0.9503337$.

That means the vaccine BNT162b2 can reduce risk of having COVID 19 by around 95\% compare to placebo.

We will utilize these approach in the non-parametric bootstrapping approach.

2) Second, we can further reduce the problem to a one sample problem by conditioning on the number of total cases and making more assumptions:

Let $S$ denote the total cases in two groups where $S=X+Y$ and let $T$ denotes the number of cases in vaccine group from $S$ total cases. 

Let $V$ denote the event that "being assigned to BNT162b2 group" and $I$ denote the event that "infections".

Then, $P(V)$, the probability of subjects being assigned to the BNT162b2 group, is $\frac{n_1}{n_1+n_2}$. Then, by the rule of complement, $P(V^c)$, the probability of subjects being assigned to the placebo group, is $\frac{n_2}{n_1+n_2}$.

Then, by distributive laws, $P(I)$, the probability of infections, is:
\begin{align*}
P(I) &= P(I\cap (V\cup V^c))=P((I\cap V)\cup(I\cap V^c))\\
&=P(I \cap V) + P(I \cap V^c) -P((I \cap V)\cap(I \cap V^c))\qquad\mbox{by addition rule}\\
&= P(I \cap V) + P(I^c \cap V^c)\qquad\mbox{since complement sets are disjointed}\\
&=P(I|V) P(V) + P(I|V^c) P(V^c) \qquad\mbox{by definition of conditional probability} \\
&= \pi_v \cdot \frac{n_1}{n_1+n_2} + \pi_p \cdot \frac{n_2}{n_1+n_2} \\
&= \frac{n_1 \pi_v + n_2 \pi_p}{n_1 + n_2}
\end{align*}
Let $\pi$ denote the probability of a subject with the BNT162b2 vaccine but in the infection group, that is $\pi = P(V|I)=P(\text{BNT162b2 | COVID 19})$.

Since participants underwent randomization in the paper, every case is independent of each other. With independence and a fixed probability of $\pi$, a reasonable model that we can use for $T$ is $T \sim Binom(s, \pi)$ where $\pi = P(V|I)$. Now, let's find $\pi$:

By Bayes’ rule, we know that $P(V|I) = \frac{P(I|V)P(V)}{P(I)}$

where $P(I|V)$ represents the probability of infections given that being assigned to BNT162b2 group, which is $\pi_v$ and $P(I|V^c)$ represents the probability of infections given that being assigned to placebo group, which is $\pi_p$.

Then, by definition of conditional probability, we can calculate $\pi = P(V|I)$ as follows:
$\pi = P(V|I) = \frac{P(I|V)P(V)}{P(I)} = \frac{\pi_v \cdot \frac{n_1}{n_1+n_2}}{\frac{n_1 \pi_v + n_2 \pi_p}{n_1 + n_2}} = \frac{n_1\pi_v}{n_1+n_2} \cdot \frac{n_1+n_2}{n_1 \pi_v + n_2 \pi_p} = \frac{n_1\pi_v}{n_1 \pi_v + n_2 \pi_p}$

By the table, we know that $n_1=17411$ and $n_2=17511$. Since $n_1 \approx n2$, we can say that the randomization is 1:1 and further approximate $\pi$ as $\pi = \frac{n_1\pi_v}{n_1\pi_v+n_2\pi_p}= \frac{n_1\pi_v}{n_1\pi_v+n_1\pi_p}= \frac{n_1\pi_v}{n_1(\pi_v+\pi_p)}= \frac{\pi_v}{\pi_v+\pi_p}$.

Then, the parameter of interest, which is the vaccine efficacy represented by $\pi$ is:
$\psi = 1 - \frac{\pi_v}{\pi_p} = 1 - \frac{\pi}{1-\pi} = \frac{1-2\pi}{1-\pi}$.

At the same time, we also know that $\pi = \frac{1 - \psi}{2 - \psi}$.

- (we can put these steps in the appendix)
The parameter of interest is vaccine efficacy $\psi$, that is 

\begin{align*}
\psi &= \frac{\pi_p - \pi_v}{\pi_p} \\
&= 1 - \frac{\pi_v}{\pi_p} \\
\end{align*}

Then, we can find $\frac{\pi_v}{\pi_p}$ by using $\pi$

\begin{align*}
\pi &= \frac{\pi_v}{\pi_v+\pi_p} \\
\pi(\pi_v+\pi_p) &= \pi_v \\
\pi \pi_v + \pi \pi_p &= \pi_v \\
\pi \pi_p &= \pi_v - \pi \pi_v \\
\pi \pi_p &= \pi_v (1 - \pi) \\
\frac{\pi_v}{\pi_p} &= \frac{\pi}{1-\pi}
\end{align*}
- (we can put these steps in the appendix)

Based on the data from the table, we can easily calculated that $s=x+y=8+162=170$ and $\pi=\frac{\pi_v}{\pi_v+\pi_p}=0.04731632$. Then, $T \sim Binom(s=170, \pi=0.04731632)$ where $\pi_v$ and $\pi_p$ are the probabilities of an infection in the vaccine and placebo groups respectively.

We will utilize these approach in the Maximum Likelihood and Bayesian approach.

### Hypothesis of Interest

(To do hypothesis test for inference, we first need to make certain assumptions to make our test valid.) 

The FDA requires at least 30% efficacy for a new therapy to be approved. Let $\psi_0$ represent the true value of $\psi$, which is the vaccine efficacy for BNT162b2.

Then, our null hypothesis is: "The vaccine efficacy of BNT162b2 is 30%", which is $H_0: \psi_0=0.3$.

Our alternative hypothesis is: "The vaccine efficacy of BNT162b2 over 30%", which is $H_1: \psi_0>0.3$.


## Maximum Likelihood 

We used the one sample approach to solve the question. Let's find the Maximum Likelihood Estimator first.

Based on 4.1.2, we have know that $T$, the number of cases in vaccine group from total cases has the distribution $T \sim Binom(170, \pi_0)$ where $\pi_0 = \frac{1 - \psi_0}{2 - \psi_0}$.

We know that for Binomial distribution, the PDF is $f(t)=\binom{170}{t} \cdot \pi_0^{t} \cdot (1 - \pi_0)^{170 - t}$ for $0<\pi_0<1$. Then, replace $\pi_0$ as $\pi_0 = \frac{1 - \psi_0}{2 - \psi_0}$, we got the PDF of $T$ represented by $\psi_0$, which is $f(t)=\binom{170}{t} \cdot (\frac{1 - \psi}{2 - \psi})^{t} \cdot (1 - \frac{1 - \psi}{2 - \psi})^{170 - t}$ for $0<\frac{1 - \psi_0}{2 - \psi_0}<1$, which is $0<\psi<1$.

Then, indexed by a parameter $\psi$ with true value $\psi_0$ the likelihood function is:
\begin{align*}
L(\psi) &= \binom{170}{t} \cdot \left(\frac{1 - \psi}{2 - \psi}\right)^{t} \cdot \left(1 - \frac{1 - \psi}{2 - \psi}\right)^{170 - t} \\
&= \binom{170}{t} \cdot \left(\frac{1 - \psi}{2 - \psi}\right)^{t} \cdot \left(\frac{1}{2 - \psi}\right)^{170 - t}\qquad\mbox{for $0<\psi<1$}
\end{align*}
Since the $\psi$ that maximizes $L(\psi)$ also maximizes $\ell(\psi) = lnL(\psi)$, and $\ell(\psi)$ is much easier to work with, let’s
calculate the $\ell(\psi)$.
\begin{align*}
\ell(\psi) &= ln\left(\binom{170}{t} \cdot (\frac{1 - \psi}{2 - \psi})^{t} \cdot (\frac{1}{2 - \psi})^{170 - t}\right) \\
&= ln\left(\binom{170}{t}\right) + t \cdot (ln(1 - \psi) - ln(2 - \psi)) + (t - 170) \cdot ln(2 - \psi)\qquad\mbox{for $0<\psi_0<1$}
\end{align*}
Here, to maximize $\psi$, we need to take the derivative of $\ell(\psi)$ with respect to $\psi$ and set it to 0 to find the critical point. We also need to take the second derivative of $\ell(\psi)$ with respect to $\psi$ to prove that the critical point we get from the first derivative is the global maximum. If we only have one critical point and the second
derivative of $\ell(\psi)$ with respect to $\psi$ on that critical point is negative, that means the function $\ell(\psi)$ is always concave down and the critical point has the value of the global maximum for the function $\ell(\psi)$.

Let's start with the first derivative:
\begin{align*}
\frac{d}{d \psi} \ell(\psi) &= \frac{d}{d \psi} \left(ln\left(\binom{170}{t}\right) + t \cdot (ln(1 - \psi) - ln(2 - \psi)) + (t - 170) \cdot ln(2 - \psi)\right)=0 \\
&= -\frac{t}{1 - \psi} + \frac{t}{2 - \psi} - \frac{t - 170}{2 - \psi}=0 \\
&= -\frac{t}{1 - \psi} + \frac{170}{2 - \psi} = 0 \\
170 - 170 \cdot \psi &= 2t - t \cdot \psi \\
(170 - t) \cdot \psi &= 170 - 2t \\
\psi &= \frac{170 - 2t}{170 - t}
\end{align*}
Thus, $\frac{170 - 2t}{170 - t}$ is a candidate for MLE for $\psi_0$. To check whether this candidate is a maximum value, we need to calculate the second derivative test:
\begin{align*}
\frac{d^2}{d \psi^2} \ell(\psi) = \ell''(\psi) &= \frac{d}{d \psi} \left(-\frac{t}{1 - \psi} + \frac{170}{2 - \psi}\right) \\
&= \frac{-t}{(1 - \psi)^{2}} + \frac{170}{(2 - \psi)^{2}} \\
&= \frac{-t\cdot (2 - \psi)^{2} + 170 \cdot (1 - \psi)^{2}}{(1 - \psi)^{2} \cdot (2 - \psi)^{2}}
\end{align*}
From Table 1, given that 8 cases in Vaccine group, we know that $t=8$, then $\psi = \frac{170 - 2\cdot 8}{170 - \cdot 8}=0.9506173$. Then, $\ell''(\psi)=\frac{-8\cdot (2 - 0.9506173)^{2} + 170 \cdot (1 - 0.9506173)^{2}}{(1 - 0.9506173)^{2} \cdot (2 - 0.9506173)^{2}}=-3126.12565702$.

Since we only have one critical point and the second derivative of $\ell(\psi)$ with respect to $\psi$ on that critical point is negative, we can say that the function $\ell(\psi)$ is always concave down and the critical point, $\psi = \frac{170 - 2t}{170 - t}=0.9506173$ is the global maximum of $\ell(\psi)$.

Therefore, $\widehat{\psi^{mle}} = 0.9506173$ is a MLE for $\psi$. 

Based on the given information in Chapter 25, we have the estimated variance of the MLE of $\psi$ is $-\frac{1}{\ell''(\widehat{\psi^{mle}})}$

\begin{align*}
Var[\widehat{\psi^{mle}}] &= -\frac{1}{\ell''(\widehat{\psi^{mle}})} \\
&= -\frac{1}{\frac{-8}{(1 - 0.9506173)^{2}} + \frac{170}{(2 - 0.9506173)^{2}}} \\
&= -\frac{1}{\frac{-8}{0.002438651} + \frac{170}{1.101204}} \\
&= -\frac{1}{-3280.502 + 154.3765} \\
&= 0.0003198848 \\
SD[\widehat{\psi^{mle}}] &= \sqrt{Var[\widehat{\psi^{mle}}]} \\
&= 0.01788532 \\
\end{align*}

Then, to find the confidence interval, we have: 

\begin{align*}
\widehat{\psi^{mle}} &\pm Z \cdot \widehat{SE} \\
\widehat{\psi^{mle}} &\pm Z \cdot \sqrt{-\frac{1}{\ell''(\widehat{\psi^{mle}})}} \\
0.9506173 &\pm 1.96 \cdot 0.01788532 \\
[0.9155621, & \: 0.9856725] \\
\end{align*}

Because the null hypothesis is $\psi^{null} = \frac{1 - 2\pi}{1 - \pi} = 0.3$ and $\widehat{\psi^{mle}} = 0.9506173$.

\begin{align*}
W &= 2[\ell(\widehat{\psi^{mle}}) - \ell(\psi^{null})] \\
&= 2[ln\left(\binom{170}{t}\right) + t \cdot (ln(1 - \widehat{\psi^{mle}}) - ln(2 - \widehat{\psi^{mle}})) + (t - 170) \cdot ln(2 - \widehat{\psi^{mle}}) - \\
& \quad (ln\left(\binom{170}{t}\right) + t \cdot (ln(1 - \psi^{null}) - ln(2 - \psi^{null})) + (t - 170) \cdot ln(2 - \psi^{null}))] \\
&= 2[t \cdot ((ln(1 - \widehat{\psi^{mle}}) - ln(2 - \widehat{\psi^{mle}})) - (ln(1 - \psi^{null}) + ln(2 - \psi^{null}))) \\
& \quad + (t - 170) \cdot (ln(2 - \widehat{\psi^{mle}})-ln(2 - \psi^{null}))] \\
&= 2[t \cdot (ln(0.0493827) - ln(1.0493827) - ln(0.7) - ln(1.7)) + (t - 170) \cdot (ln(1.0493827)-ln(1.7))] \\
&= 2[t \cdot -3.230311 + (t - 170) \cdot -0.4824262] \\
\text{Given $t = 8$: } &= 2[8 \cdot -3.230311 + (-164) \cdot -0.4824262] \\
&= 2[8 \cdot -3.230311 + (-164) \cdot -0.4824262] \\
&= 106.5508 \\
\end{align*}

```{r label="likelihood", include = FALSE, cache = TRUE}
pchisq(q = 106.5508, df = 1, lower.tail = F)
pchisq(q = 106.5508, df = 1, lower.tail = T)
```

Because at $t = 8$, we have $W = 106.5508$, and because the P-value for $P(w < 106.5508) = 5.584252e-25$, which is very small. So, we have enough evidence to reject the null hypothesis and we have $P(\psi > 0.3) = 1$. Therefore, we can conclude that the vaccine efficacy is greater than 30\%.

## Bayesian

Let $X$ denote the number of infection cases among $n_1 = 17411$ subjects assigned to the BNT162b2 group

Let $Y$ denote the number of infection cases among $n_2 = 17511$ subjects assigned to the Placebo group.

Let $\pi_v$ and $\pi_p$ denote the probabilities of an infection in the vaccine and placebo groups respectively.

Therefore, we have $X \sim Binom(\pi_v, n_1 = 17411)$ and $Y \sim Binom(\pi_p, n_2 = 17511)$. 

Assuming the trial will run until $X+Y=s$ infections observed

Then, we can let T denote the number of cases in vaccine group from s=170 cases.

The distribution of T is well approximated by a bionomial for the vaccine study:

$$T \sim Binom(s,\pi)$$
where $\pi =$ P(vaccine|infection)

By Bayes'rule, we can write $\pi$ in terms of the probabilities $\pi_v$ and $\pi_p$ as follows:

$$\pi = \frac{n_1\pi_v}{n_1\pi_v+n_2\pi_p}$$
We know that $n_1=17411$ and $n_2=17511$

so $n_1 \approx n2$, we can say that the randomization is 1:1.

We can simplify $\pi$ as follows

\begin{align*}
\pi &= \frac{n_1\pi_v}{n_1\pi_v+n_2\pi_p} \\
&= \frac{n_1\pi_v}{n_1\pi_v+n_1\pi_p} \\
&= \frac{n_1\pi_v}{n_1(\pi_v+\pi_p)} \\
&= \frac{\pi_v}{\pi_v+\pi_p}
\end{align*}

Since $n_1$ and $n_2$ are large and the event rates are small, we can say

$$X \approx Pois(n_1\pi_v)$$
is independent of

$$Y \approx Pois(n_2\pi_p)$$

Then

\begin{align*}
P(T=t) &= P(X=t|X+Y=s) \ t=0,1,\dots, s \\
&= \frac{P(X=t \cap X+Y=s)}{P(X+Y=s)} \\
&= \frac{P(X=t \times Y=s-t)}{P(X+Y=s)} \text{ X and Y are indepedent} \\
&= \binom{s}{n} \pi^t (1-\pi)^{s-t} \text{where } \pi = \frac{\pi_v}{\pi_v+\pi_p}
\end{align*}

so we can have that

$$T \sim Binom(s=170, \pi=\frac{\pi_v}{\pi_v+\pi_p})$$

where $\pi_v$ and $\pi_p$ are the probabilities of an infection in the vaccine and placebo groups respectively.

The parameter of interest is vaccine efficacy $\psi$, that is 

\begin{align*}
\psi &= \frac{\pi_p - \pi_v}{\pi_p} \\
&= 1 - \frac{\pi_v}{\pi_p} \\
\end{align*}

Then, we can find $\frac{\pi_v}{\pi_p}$ by using $\pi$

\begin{align*}
\pi &= \frac{\pi_v}{\pi_v+\pi_p} \\
\pi(\pi_v+\pi_p) &= \pi_v \\
\pi \pi_v + \pi \pi_p &= \pi_v \\
\pi \pi_p &= \pi_v - \pi \pi_v \\
\pi \pi_p &= \pi_v (1 - \pi) \\
\frac{\pi_v}{\pi_p} &= \frac{\pi}{1-\pi}
\end{align*}

Then, we can find $\psi$

$\psi = 1 - \frac{\pi_v}{\pi_p} = 1 - \frac{\pi}{1-\pi} = \frac{1-2\pi}{1-\pi}$


```{r, warning=F}
library(LearnBayes)

n <- 170
t <- 8

prior_alpha <- 0.700102
prior_beta <- 1

prior_median_pi <- qbeta(p=0.5, shape1=prior_alpha, shape2=prior_beta)
prior_median_pi
prior_0.05_pi <- qbeta(p=0.05, shape1=prior_alpha, shape2=prior_beta)
prior_0.05_pi

CI_pi_prior <- qbeta(p = c(0.025,0.975), shape1=prior_alpha, shape2=prior_beta)
CI_pi_lower <- CI_pi_prior[1]
CI_pi_upper <- CI_pi_prior[2]

prior_psi <- beta.select(quantile1 = list(p=0.5, x=0.4087779),
                             quantile2 = list(p=0.95, x=0.9859487))
prior_psi_alpha <- prior_psi[1]
prior_psi_alpha
prior_psi_beta <- prior_psi[2]
prior_psi_beta 

posterior_median_pi <- qbeta(p=0.5, shape1=t+prior_alpha, shape2=n-t+prior_beta)
posterior_median_pi
posterior_0.05_pi <- qbeta(p=0.05, shape1=t+prior_alpha, shape2=n-t+prior_beta)
posterior_0.05_pi

posterior_psi <- beta.select(quantile1 = list(p=0.5, x=0.94855003),
                         quantile2 = list(p=0.95, x=0.9727862))

posterior_psi_alpha <- posterior_psi[1]
posterior_psi_alpha
posterior_psi_beta <- posterior_psi[2]
posterior_psi_beta 
```

\begin{align*}
\psi &= \frac{1-2\pi}{1-\pi} \\
\psi (1-\pi) &= 1-2\pi \\
\psi - \psi \pi &= 1-2\pi \\
2\pi - \psi \pi &= 1 - \psi \\
\pi(2-\psi)  &= 1 - \psi \\
\pi &= \frac{1 - \psi}{2-\psi}
\end{align*}

Then, we can find the prior distribution of $\psi$

By using R, the prior median of $\pi$ is `r prior_median_pi`, and the prior 5 percentile of $\pi$ is `r prior_0.05_pi`.

Then, we can calculate the prior median of $\psi$ and prior 95 percentile of $\psi$

\begin{align*}
P(\pi > 0.3715522|t) &= 0.5 \\
P(\frac{1-\psi}{2-\psi} < 0.3715522|t) &= 0.5 \\
P(1-\psi < 0.3715522(2-\psi)|t) &= 0.5 \\
P(1-\psi < 0.7431044 - 0.3715522\psi|t) &= 0.5 \\
P(-0.6284478\psi < -0.2568956|t) &= 0.5 \\
P(\psi < 0.4087779|t) &= 0.5 \\
\end{align*}

\begin{align*}
P(\pi > 0.01385659|t) &= 0.95 \\
P(\frac{1-\psi}{2-\psi} < 0.01385659|t) &= 0.95 \\
P(1-\psi < 0.01385659(2-\psi)|t) &= 0.95 \\
P(1-\psi < 0.02771318 - 0.01385659\psi|t) &= 0.95 \\
P(-0.9861434\psi < -0.9722868|t) &= 0.95 \\
P(\psi < 0.9859487|t) &= 0.95 \\
\end{align*}

Thus, we have that the prior median of $\psi$ is 0.4087779, and the prior 95 percentile of $\psi$ is 0.9859487.

Therefore, we can say that the prior distribution of $\psi$ is: 

$g(\psi_0)$ = Beta(`r prior_psi_alpha`, `r prior_psi_beta`)

Then, we can find the posterior distribution of $\psi$

By using R, the posterior median of $\pi$ is `r posterior_median_pi`, and the posterior 5 percentile of $\pi$ is `r posterior_0.05_pi`.

\begin{align*}
P(\pi > 0.04893234|t) &= 0.5 \\
P(\frac{1-\psi}{2-\psi} < 0.04893234|t) &= 0.5 \\
P(1-\psi < 0.04893234(2-\psi)|t) &= 0.5 \\
P(1-\psi < 0.09786468 - 0.04893234\psi|t) &= 0.5 \\
P(-0.9510677\psi < -0.9021353|t) &= 0.5 \\
P(\psi < 0.94855003|t) &= 0.5 \\
\end{align*}

\begin{align*}
P(\pi > 0.02649278|t) &= 0.95 \\
P(\frac{1-\psi}{2-\psi} < 0.02649278|t) &= 0.95 \\
P(1-\psi < 0.02649278(2-\psi)|t) &= 0.95 \\
P(1-\psi < 0.05298556 - 0.02649278\psi|t) &= 0.95 \\
P(-0.9735072\psi < -0.9470144|t) &= 0.95 \\
P(\psi < 0.9727862|t) &= 0.95 \\
\end{align*}

Thus, we have that the posterior median of $\psi$ is 0.94855003, and the posterior 95 percentile of $\psi$ is 0.9727862.

Therefore, we can say that the posterior distribution of $\psi$ is:

$h(\psi_0|t)=$ Beta(`r posterior_psi_alpha`, `r posterior_psi_beta`)


```{r}
library(ggplot2)

ggplot()+
  stat_function(fun = dbeta,
                args = list(shape1 = prior_psi_alpha,
                            shape2 = prior_psi_beta))+
  stat_function(fun = dbeta,
                args = list(shape1 = posterior_psi_alpha,
                            shape2 = posterior_psi_beta),
                color='blue')+
  labs(title = "Prior(black) and Posterior(blue) for Vaccine Data",
       ylab = "Density",
       xlab = expression(psi))
```

The posterior distribution is skewed, so it is better to use HPDI for an interval estimate.

```{r}
CI_psi <- qbeta(p = c(0.025,0.975), shape1=posterior_psi_alpha, shape2=posterior_psi_beta)
CI_psi
CI_psi_lower <- CI_psi[1]
CI_psi_upper <- CI_psi[2]

library(HDInterval)
credible <- hdi(qbeta, credMass = 0.95, shape1=posterior_psi_alpha, shape2=posterior_psi_beta)
credible
credible_lower <- credible[1]
credible_upper <- credible[2]
```

The 95\% Bayesian credible interval is [`r CI_psi_lower`, `r CI_psi_upper`], which indicate that there is a 95% probability that the true value of $\psi_0$ falls within the interval [`r CI_psi_lower`, `r CI_psi_upper`]. We are 95% confident that the true vaccine efficacy against Covid 19 is between approximately 90.59% and 97.63%.

The 95\% Bayesian HPDI interval is [`r credible_lower`, `r credible_upper`]. For $\psi$ represents the vaccine efficacy against Covid 19, the middle 95% of the posterior distribution of vaccine efficacy resulting from the prior is the range 91.05% to 97.92%. There is a 95% probability that $\psi_0$ in between 91.05% and 97.92%. .

Calculate posterior probability $P(\psi_0 > 0.3|t)$

```{r}
post_prob <- pbeta(q=0.3, shape1=posterior_psi_alpha, shape2=posterior_psi_beta, lower.tail = F)
post_prob
smaller_0.3 <- pbeta(q=0.3, shape1=posterior_psi_alpha, shape2=posterior_psi_beta)
smaller_0.3
```

The posterior probability $P(\psi_0 > 0.3|t)=1$, which indicates that approximately 100% of chance that the true vaccine efficacy against Covid 19 is greater than 30%. We can reject the null hypothesis that the vaccine efficacy is 30%. The evidence suggests that vaccine efficacy is greater than 30%.

### The burning question is whether the prior used by Pfizer was non-informativeor was it overly optimistic? 

We choose Beta(0.700102, 1) as the prior distribution of $\pi$. The prior distribution is chosen such that the mean is equal to 0.0471 ($\frac{8}{170}$) corresponding to expected vaccine efficacy 30%. The mean is much lower than vaccine efficacy = 30%, which means that it assigns a lower probability to values indicating higher vaccine efficacy, so the prior used by Pfizer was overly optimistic. Pfizer expressed a relatively high expectation for the vaccine's efficacy before observing the actual data.

### Method 2 (interval)

The parameter of interest is vaccine efficacy $\psi$, that is 

\begin{align*}
\psi &= \frac{\pi_p - \pi_v}{\pi_p} \\
&= 1 - \frac{\pi_v}{\pi_p} \\
\end{align*}

Then, we can find $\frac{\pi_v}{\pi_p}$ by using $\pi$

\begin{align*}
\pi &= \frac{\pi_v}{\pi_v+\pi_p} \\
\pi(\pi_v+\pi_p) &= \pi_v \\
\pi \pi_v + \pi \pi_p &= \pi_v \\
\pi \pi_p &= \pi_v - \pi \pi_v \\
\pi \pi_p &= \pi_v (1 - \pi) \\
\frac{\pi_v}{\pi_p} &= \frac{\pi}{1-\pi}
\end{align*}

Then, we can find $\psi$

$\psi = 1 - \frac{\pi_v}{\pi_p} = 1 - \frac{\pi}{1-\pi} = \frac{1-2\pi}{1-\pi}$

Then, we can find the expression for $\pi$ with $\psi$

\begin{align*}
\psi &= \frac{1-2\pi}{1-\pi} \\
\psi (1-\pi) &= 1-2\pi \\
\psi - \psi \pi &= 1-2\pi \\
2\pi - \psi \pi &= 1 - \psi \\
\pi(2-\psi)  &= 1 - \psi \\
\pi &= \frac{1 - \psi}{2-\psi}
\end{align*}


### Prior Elicitation

Under the null hypothesis, the prior value of vaccine efficacy($\psi$) is 30%, so we can have that

$\pi = \frac{1-\psi}{2-\psi} = \frac{1-0.3}{2-0.3} = 0.4118$

We know that $\pi$ follows beta distribution, so the expected value of $\pi$ is $\frac{\alpha}{\alpha+\beta}$

Assume that prior is centered closer to 0, indicating a prior belief that $\alpha$ and $\beta$ is more likely to have lower values, so we choose $\beta = 1$ and a smaller $alpha$

\begin{align*}
\frac{\alpha}{\alpha+1} &= 0.4118 \\
\alpha &= 0.4118(\alpha+1) \\
\alpha &= 0.4118\alpha + 0.4118\\
0.5882\alpha &= 0.4118 \\
\alpha &= 0.700102
\end{align*}

Therefore, the prior of $\pi$ is

$$g(\pi_0) = Beta(0.700102, 1)$$

We also choose a non informative uniform prior distribution of $\pi$

$$g(\pi_0) = Beta(1,1)$$

### Calculating the posterior distribution

For prior Beta(0.700102, 1)

$T \sim Binom(s=170, \pi=\frac{\pi_v}{\pi_v+\pi_p})$

$h(\pi_0|t) = Beta(0.700102+8, 170-8+1) = Beta(8.700102, 163)$

```{r}
library(ggplot2)

ggplot()+
  geom_function(fun = dbeta,
                args = list(shape1 = 0.700102,
                            shape2 = 1),
                mapping = aes(color = "prior")) +
  geom_function(fun = dbeta,
                args = list(shape1 = 0.700102+8,
                            shape2 = 170-8+1),
                mapping = aes(color = "posterior")) +
  labs(title = "Beta(0.700102, 1) Prior and Beta(8.700102, 163)Posterior
       for Vaccine Data",
       y = "PDF",
       x = expression(pi),
       color = "Distribution")
```

```{r}
interval <- qbeta(p = c(0.025,0.975), shape1=8.700102, shape2=163)
interval
```

The 2.5 percentile of posterior distribution of $\pi$ is 0.02319402, and the 97.5% percentile of posterior distribution of $\pi$ is 0.08799074.

We can calculate the 95\% Bayesian credible interval, that is 

$[\frac{1-2 \times 0.08799074}{1-0.08799074},\frac{1-2 \times 0.02319402}{1-0.02319402}] = [0.9035199, 0.9762552]$

For prior Beta(1, 1)

$T \sim Binom(s=170, \pi=\frac{\pi_v}{\pi_v+\pi_p})$

$h(\pi_0|t) = Beta(1+8, 170-8+1) = Beta(9, 163)$

```{r}
library(ggplot2)

ggplot()+
  geom_function(fun = dbeta,
                args = list(shape1 = 1,
                            shape2 = 1),
                mapping = aes(color = "prior")) +
  geom_function(fun = dbeta,
                args = list(shape1 = 1+8,
                            shape2 = 170-8+1),
                mapping = aes(color = "posterior")) +
  labs(title = "Beta(1, 1) Prior and Beta(9, 163) Posterior for Vaccine Data",
       y = "PDF",
       x = expression(pi),
       color = "Distribution")
```

```{r}
interval_uniform <- qbeta(p = c(0.025,0.975), shape1=9, shape2=163)
interval_uniform 
```

The 2.5 percentile of posterior distribution of $\pi$ is 0.02434818, and the 97.5% percentile of posterior distribution of $\pi$ is 0.092214.

We can calculate the 95\% Bayesian credible interval, that is 

$[\frac{1-2 \times 0.09009896}{1-0.09009896},\frac{1-2 \times 0.02434585}{1-0.02434585}] = [0.9009794,0.9750466]$

### Calculate a Bayesian P-value

We can calculate posterior probability, that is

$$P(\psi_0 \leq 0.3|t) = P(\frac{1-2\pi}{1-\pi} \leq 0.3|t) = P(\pi \geq 0.4117647|t)$$

```{r}
post_prob_pi <- pbeta(q=0.4117647, shape1=8.700102, shape2=163, lower.tail = F)
post_prob_pi
```

The posterior probability $P(\psi_0 \leq 0.3|t)=1.96 \times 10^{-28}$, which is very small, so $P(\psi_0 > 0.3|t)$ is approximately 1, which means that the probability that the true vaccine efficacy against Covid 19 is greater than 30% is about 100%. The evidence suggests that vaccine efficacy is greater than 30%.

## Non-parametric Bootstrap 
```{r label="bootstrap", include = FALSE, cache = TRUE, warning=F}
set.seed(666)

case_vaccine <- 8
total_vaccine <- 17411
case_placebo <- 162
total_placebo <- 17511
total_case <- case_vaccine + case_placebo

pi_v <- case_vaccine / total_vaccine
pi_p <- case_placebo / total_placebo
psi_hat <- (pi_p - pi_v) / pi_p
pi_hat <- pi_v / (pi_p + pi_v)

B <- 10000

a = 0.05
z <- qnorm(p = 1 - a/2)
h_0_psi <- 0.3
```

To use the non parametric bootstrapping to construct a 95% bootstrap confidence interval for the BNT162b2 vaccine efficacy $\psi$, we do not need to make any assumption about the probability distribution of the sample. We can directly generate 10000 $\pi_v$ and $\pi_p$ to calculate the exact $\psi$ for each generated sample and do the inference. Here, we assume that the population distribution for the infection in the vaccine group is a discrete probability distribution that gives probability $\frac{1}{170}$ (since we have a total number of 170 infections) to each observed participants who is vaccinated the BNT162b2 vaccine but is infected by COVID 19. We assume that the population distribution for the infection in the placebo group is a discrete probability distribution that gives probability $\frac{1}{170}$ to each observed participants who is not vaccinate the BNT162b2 vaccine and is infected by COVID 19.

After generating 10000 bootstrapped estimates of $\psi_0$, we can see that from the histogram the bootstrapped sampling distribution is skewed to the right and the sample data points on the quantile quantile plot do not fit the line very well. Therefore, a normal distribution is not a good fit for our bootstrapped samples of $\psi_0$. Thus, we cannot construct the standard bootstrap 95% interval. Instead, here, we construct the 95% bootstrap percentile interval with the simple percentile method because it doesn't need any assumption about the distribution of $\psi_0$.

```{r label="interval", include = FALSE, cache = TRUE, warning = F}
non_bootstrap_samples <- lapply(1:B, function(X) {
  # Resample from the combined dataset
  bootstrap_data <- sample(c(rep("vaccine", case_vaccine), 
                             rep("placebo", case_placebo)), 
                             total_case, replace = TRUE)
  bootstrap_case_vaccine <- sum(bootstrap_data == "vaccine")
  bootstrap_case_placebo <- sum(bootstrap_data == "placebo")
  bootstrap_pi_v <- bootstrap_case_vaccine / total_vaccine
  bootstrap_pi_p <- bootstrap_case_placebo / total_placebo

  bootstrap_psi_hat <- (bootstrap_pi_p - bootstrap_pi_v) / bootstrap_pi_p
  
  data.frame(sample_psi_hat = bootstrap_psi_hat)
})

non_sample_summary <- do.call(rbind, non_bootstrap_samples)

# Calculate the binwidth using the method we have learned in class:
# Here, n = B = 10000
# log2(10000) + 1 = 14.28771
# Here, I took a number of 15.
# Now, I need the range from my data.
# para_sample_summary$sample_psi_hat %>% range()
# From the summary, I got the range for the data is: 1.0000000 - 0.8659009 = 0.1340991
# Thus, here, I resize the binwidth into 0.1340991/15 and breaks the bar by (0.8659009, 1.0000000, 0.1340991/15)

p1 <- ggplot(data = non_sample_summary,
       mapping = aes(x = sample_psi_hat)) + 
  geom_histogram(binwidth = 0.1258278/15,
                 breaks = seq(0.8741722, 1.0000000, 0.1258278/15), 
                 alpha = 0.8,
                 color = 'gray') +
  labs(title =  expression(paste("The Bootstrapped \nSampling Distribution of ", widehat(psi))),
       subtitle = "n=10000",
       x = expression(widehat(psi)))

p2 <- ggplot(data = non_sample_summary, 
             mapping = aes(sample = sample_psi_hat)) +
  stat_qq(distribution = qnorm) + 
  stat_qq_line(distribution = qnorm) + 
  labs(title = "The Normal Probability Plot of \nthe Bootstrapped Distribution",
       subtitle = "n=10000",
       x = "theoretical",
       y = "sample")
p1+p2

lower_ci <- quantile(non_sample_summary$sample_psi_hat, a/2)
upper_ci <- quantile(non_sample_summary$sample_psi_hat, 1 - a/2)

# Perform hypothesis testing
non_p_value <- sum(non_sample_summary$sample_psi_hat > h_0_psi) / B

lower_ci
upper_ci
non_p_value
```

```{r echo = FALSE, warning=F}
p1+p2
```

Calculated by the R, the 95% simple bootstrap confidence interval for the true value of the vaccine efficacy $\psi_0$ is (`r lower_ci`, `r upper_ci`). That is, we are 95% confident that the true value of the vaccine efficacy $\psi_0$ is between `r lower_ci` and `r upper_ci`. 

The p-value is defined by presuming $\psi=0.3$ is true, of observing our test statistic or a value that is larger than $\psi=0.3$. Since all $\psi$ in our generate sample are greater or equal to 0.8659009, all of the generated samples' test statistic are greater than 0.3, which in the null hypothesis. Therefore, the p-value is 1. That is we are 100% confident that our test statistics contradicts the null hypothesis. 

# Results

\begin{table}[h]
\centering
\caption{Comparsion Table (ConfI: 95\% Confidence Interval, CredI: 95\% Credible Interval)}
\begin{tabular}{|c|c|c|c|}\hline
Approach & Lower Bound & Upper Bound & p-value \\ \hline
Maximum Likelihood (ConfI) & 0.9155621 & 0.9856725 & 1  \\ \hline
Bayesian prior Beta(0.700102, 1) (CredI) &  0.9035199 & 0.9762552 & 1 \\ \hline
Bayesian Prior Beta(1, 1) (CredI) & 0.9009794 & 0.9750466 & 1 \\ \hline
Non-parametric Bootstrap (ConfI) & 0.9097410 & 0.9819328 & 1\\ \hline
\end{tabular}
\end{table}

  In this study, we have only 0.05\% of cases in the treatment group while we have 0.93\% of cases in the control group. In the Likelihood method, we first analyze the efficiency of BNT162b2 by comparing the parameters of control group and treatment group. We found $\psi = 0.9503337$, shows that we have 95\% efficiency of the treatment group, which pass the rule of at least 30\% efficacy for a new therapy to be approved. Moreover, we combined the results of both groups together to find the combined maximum likelihood estimator, which is 0.9506173. And then, use this estimator, we build a 95\% confidence interval: [0.9155621, 0.9856725]. We also calculate at $t = 8$, $W = 106.5508$, and by this, we have $P(W >106.5508) = P(\psi > 0.3) = 1$
  
  Then, in the Bayesian analysis, based on the given information and the prior distribution: $Beta(0.700102, 1)$, we then can calculate the posterior distribution: $Beta(8.700102, 163)$. Also, the 95\% Bayesian credible interval is [0.9035199, 0.9762552]. Besides, we consider a non informative uniform prior distribution of $\pi$: $Beta(1,1)$. The posterior distribution for this non informative uniform prior distribution is $Beta(9,163)$ and the 95\% Bayesian credible interval is [0.9009794, 0.9750466]. These results are slightly different because in Bayesian analysis, the choice of prior distribution can impact the posterior distribution and, consequently, the calculation of the confidence interval.
  
  The third method is Non-parametric Bootstrapping. By generating 10000 bootstrapped estimation of $\psi_0$, we can construct a 95% confidence interval: [0.9097410, 0.9819328].
  
  Based on table provided above, we can see the p-value for all three analysis methods are all 1.
  

# Discussion/Conclusion

  In conclusion, we con conclude that our 95\% confidence interval 


# References

Coronavirus disease (COVID-19) pandemic. (2023, May 24).                  
https://www.who.int/europe/emergencies/situations/covid-19

Hwang TJ, Rabheru K, Peisah C, Reichman W, Ikeda M. Loneliness and social isolation during the COVID-19 pandemic. Int Psychogeriatr. 2020 Oct;32(10):1217-1220. doi: 10.1017/S1041610220000988. Epub 2020 May 26. PMID: 32450943; PMCID: PMC7306546.

“Adult Vaccination - Reasons to Vaccinate.” Centers for Disease Control and Prevention, 22 Sept. 2022, www.cdc.gov/vaccines/adults/reasons-to-vaccinate.html.

# Appendix

## Code for Likelihood
```{r ref.label="likelihood", echo = TRUE, eval = F}

```

## Code for Bootstrap
```{r ref.label="bootstrap", echo = TRUE, eval = F, warning=F}

```

```{r ref.label="interval", echo = TRUE, eval = F, warning=F}

```
